---
title: "Colon Crypt Respirometry"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{colonCryptRespirometry}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

```{r setup}
library(tidyverse)
library(SeahorseR)
```

## Data import

1.  Export the data from WAVE to excel
2.  Open the excel file and select the worksheet "Rate Data"
3.  Select the data with CMD-A or Ctrl-A and copy to the clipboard (CMD-C, or Ctrl-C).
4.  Open a new plain text file in a plain text editor (RStudio or notepad) or and paste the data (CMD-V or Ctrl-V)
5.  Create a folder for rate_data and save the file with a name that specifies the experiment or batch (e.g. "2023-02-14")
6.  Normalization data should be a matrix with 12 columns 8 rows (format of a 96-well plate) and in a tab-delimited file.
7.  The bindRateData function will create a file with data from multiple runs. It can import normalized data too, but normalization can be added later, which I prefer.

```{r}

dt.raw <- bindRateData(data.dir = "rate_data",
                       norm.meth = "none",
                       out.name = "all_rate_data_raw.txt")
```

8.  remove background and any wells that need to be excluded

```{r}
dt.raw <- remove_wells(dt.raw,
                       exclude = dt.raw$well[dt.raw$wave.id%in%c("Unassigned","Background")]
                       )
```

9.  Clean it up a bit and add grouping and interval columns

This is a good time to make sure the values in wave.id look ok.

```{r}
unique(dt.raw$wave.id)

```

The periods in wave.id help with the meta data variables.

```{r}
pheno<-as.data.frame(do.call(rbind,strsplit(dt.raw$wave.id,"\\.")))
names(pheno)<-c("SampNum","mID","Tissue","GT","Age")
dt.raw<-cbind(dt.raw, pheno)
```

create variable to group by tissue, batch and mouse ID.

```{r}
dt.raw$tissueID<-paste(dt.raw$batch,
                       dt.raw$mID,
                       dt.raw$Tissue,
                       sep = "_")
```

make sure the numeric columns are numeric

```{r}
dt.raw$ocr <- as.numeric(gsub(",","",dt.raw$ocr))
dt.raw$ecar <- as.numeric(gsub(",","",dt.raw$ecar))
dt.raw$ppr <- as.numeric(gsub(",","",dt.raw$ppr))
```

add interval columns and columns to group by interval

```{r}

dt.raw$interval <- create_column_interval(dt.raw,
                                          Interval.list=list(Int0=c(1),
                                                             Int1=c(2:3),
                                                             Int2=c(4:6),
                                                             Int3=c(7:10),
                                                             Int4=c(11:14)))
dt.raw$treatment <- create_column_treatment(dt.raw,
                                            Treatment.list=list(
                                              Int1="basal",
                                              Int2="oligomycin",
                                              Int3="FCCP",
                                              Int4="Rot+AA",
                                              Int0="Start"
                                            ))
dt.raw$sample_id <- paste(dt.raw$tissueID, dt.raw$well, sep = "_")
dt.raw$tis_int <- paste(dt.raw$tissueID, dt.raw$interval, sep = "_")
dt.raw$samp_int <- paste(dt.raw$sample_id, dt.raw$interval, sep = "_")

# consolidate ages
dt.raw$AgeGrp<-ifelse(dt.raw$Age%in%c("10-11m","10-12m"),"10-12m","1-4m")

# update abbreviations
dt.raw$GT <- gsub("ASO","Thy1-haSyn",dt.raw$GT)

# consider saving the processed raw data
saveRDS(dt.raw,file = "dt.raw-processed.rds")

```

## QC

### limit to wells with good response to inhibitors

summarize by interval for each well

```{r}
dt.int <-
  dt.raw %>% 
  filter(interval!="Int0") %>% 
  calcInt(varname = "ocr.samp.int",
          groupingVar="samp_int",
          intVar="interval",
          intervals=c("Int1","Int2","Int3","Int4"),
          measure="ocr",
          calcFunctions=c("mean","min", "max", "min")) %>% 
  ungroup() %>% 
  select(sample_id,ocr.samp.int,tissueID,interval) %>% 
  distinct() %>% 
  tidyr::pivot_wider(id_cols = c(sample_id, tissueID), 
              names_from = interval, values_from = ocr.samp.int) 
head(dt.int)
```

find wells with predicted response to inhibitors (stress-pass, sp)

```{r}
dt.int.sp <- dt.int %>% 
  filter(Int3 > Int1,
         Int1 > Int2,
         Int2 > Int4) %>% 
  distinct() %>% 
  group_by(tissueID) %>% 
  mutate(nWells=n()) %>% 
  {.->> int.sp.all} %>% 
  filter(nWells > 1) %>% 
  ungroup() %>% 
  arrange(sample_id)

dt.sp <- dt.raw %>% 
  filter(sample_id %in% dt.int.sp$sample_id)

```

```{r}

wellsPerInt(dt.raw) 
wellsPerInt(dt.sp) 


```

### add column with the normalization factor

```{r}

dt.spn<-add.norm(dt.sp, 
                  norm.dat.dir = "norm-data-crypts",
                  norm.meth = "crypts",
                  batch.col = "batch")

# only include those with normalization data
dt.spn<-dt.spn[!is.na(dt.spn$crypts),]

wellsPerInt(dt.spn)

```

### detection of raw data outliers

summarize by interval

```{r}

int.spn<-dt.spn %>% 
  group_by(batch,time) %>% 
  mutate(ocr=ifelse(find_outlier(ocr,mult = 1.5),
                    NA,
                    ocr)) %>% 
  ungroup() %>% 
  {.->>dt.spno1} %>% 
  filter(interval!="Int0") %>% 
  calcInt(varname = "ocr",
          groupingVar="samp_int",
          intVar="interval",
          intervals=c("Int1","Int2","Int3","Int4"),
          measure="ocr",
          calcFunctions=c("mean","min", "max", "min")) %>% 
  select(sample_id,samp_int,interval,tissueID,
                GT,AgeGrp,batch,well,mID,crypts,ocr) %>% 
  distinct()%>% 
  ungroup() %>% 
  group_by(batch,interval) %>% 
  mutate(outlier.norm=if_else(find_outlier(ocr/crypts,mult = 1.5),
                              true = ocr/crypts,
                              false=NA,
                              missing=1),
         outlier.raw=if_else(find_outlier(ocr,mult = 1.5),
                             true = ocr,
                             false=NA,
                             missing=1)) %>%
  ungroup() %>% 
  distinct()


int.spn %>% filter(is.na(ocr))

# excluded one sample with missing data for Int 3 after removed raw outlier wells
raw.outs <- int.spn %>% 
  filter(outlier.raw==1) %>% 
  select(sample_id) %>% 
  as.character()


dt.spno1 <- dt.spno1 %>% 
  filter(!sample_id%in%raw.outs) 

int.spn <- int.spn %>% 
  filter(!sample_id%in%raw.outs)


```

wells with no raw outliers

```{r}

wellsPerInt(dt.spno1)
```

```{r,fig.cap="retained for sample size"}

ggplot(int.spn,aes(interval,ocr))+
  geom_boxplot()+
  geom_text(aes(interval,outlier.raw,label=paste(well,mID,sep = "\n")), na.rm=TRUE,size=3,nudge_x = 0, nudge_y = 0,hjust=1)+
  facet_wrap(~batch,nrow = 1)+
  ggtitle("raw outliers")

```

```{r,fig.cap="outliers in normalized data, but not in raw data"}

ggplot(int.spn,aes(interval,ocr/crypts))+
  geom_boxplot()+
  geom_text(aes(interval,outlier.norm,label=paste(well,mID,sep = "\n")), na.rm=TRUE,size=3,nudge_x = 0, nudge_y = 0,hjust=1)+
  facet_wrap(~batch,nrow = 1)

outliers1<-int.spn %>% 
  filter(!is.na(outlier.norm)) %>% 
  select(sample_id) %>% 
  distinct()
```


impute by regression crypt count for these wells

```{r}

good.crypts<-int.spn %>% 
  filter(!sample_id%in%outliers1$sample_id) %>% 
  ungroup() %>% 
  select(sample_id) %>% 
  distinct()

wells.imputed<-wellsPerInt(int.spn %>% 
                             filter(!sample_id%in%(good.crypts$sample_id)))

predict.crypts <- int.spn %>% 
  filter(interval=="Int3") %>% 
  mutate(crypts=ifelse(sample_id%in%outliers1$sample_id,NA,crypts)) %>% 
  select(sample_id,tissueID,mID,batch,well,ocr,crypts)


fit.crypts <- lm(crypts~ocr+batch, data = predict.crypts,)
summary(fit.crypts)
predict.crypts$crypts.pred<-predict(fit.crypts,predict.crypts)

dt.spn2 <- dt.spno1 %>% 
  left_join(.,predict.crypts[,c("sample_id","crypts.pred")],by="sample_id",) %>%   
  mutate(crypts.im = if_else(sample_id%in%good.crypts$sample_id,
                             true = crypts,
                             false= crypts.pred,
                             missing=crypts),
         ocr.raw=ocr,
         ecar.raw=ecar,
         ocr=ocr/crypts.im,
         ecar=ecar/crypts.im) 
```




recalculate and view outliers
```{r}
int.spn2<-dt.spn2 %>% 
  group_by(samp_int) %>% 
  filter(interval!="Int0") %>% 
  calcInt(varname = "ocr",
          groupingVar="samp_int",
          intVar="interval",
          intervals=c("Int1","Int2","Int3","Int4"),
          measure="ocr",
          calcFunctions=c("mean","min", "max", "min")) %>% 
  select(sample_id,samp_int,interval,tissueID,
                GT,AgeGrp,batch,well,mID,crypts,ocr) %>% 
  distinct()%>% 
  ungroup() %>% 
  group_by(batch,interval) %>% 
  mutate(outlier=ifelse(find_outlier(ocr,mult = 1.5),
                        ocr,NA)) %>%
  ungroup() %>% 
  distinct() 

wellsPerInt(int.spn2[is.na(int.spn2$outlier),])

```

```{r, fig.cap="outliers after imputing crypt counts"}

ggplot(int.spn2,aes(interval,ocr))+
  geom_boxplot()+
  geom_text(aes(interval,outlier,label=paste(well,mID,sep = "\n")), na.rm=TRUE,size=3,nudge_x = 0, nudge_y = 0,hjust=1)+
  facet_wrap(~batch,nrow = 1)+
  ggtitle("Outliers for removal")

```



```{r}

dt.norm<-dt.spn2 %>% 
  group_by(batch,time) %>% 
  mutate(ocr=ifelse(find_outlier(ocr,mult = 1.5),
                    NA,ocr)) %>%
  ungroup() 


int.norm <- dt.spn2 %>% 
  group_by(samp_int) %>% 
  filter(interval!="Int0") %>% 
  calcInt(varname = "ocr",
          groupingVar="samp_int",
          intVar="interval",
          intervals=c("Int1","Int2","Int3","Int4"),
          measure="ocr",
          calcFunctions=c("mean","min", "max", "min")) %>% 
  select(sample_id,samp_int,interval,ocr,tissueID,
                GT,AgeGrp,batch,ocr,well,mID) %>% 
  distinct()%>% 
  ungroup() %>% 
  group_by(batch,interval) %>% 
  mutate(ocr=ifelse(find_outlier(ocr,mult = 1.5),
                    NA,ocr)) %>%
  ungroup() %>% 
  distinct() 

```



## Calculation of respiratory states

```{r}

data.cols<-c("Basal", "Maximal",
             "Spare","ATP_linked",
             "Proton_leak","Non_Mito")

dt.states <-
  int.norm %>% 
  filter(!is.na(ocr)) %>% 
  filter(interval!="Int0") %>% 
  select(interval,ocr,
                GT,AgeGrp,sample_id,tissueID, mID,
                batch) %>% 
  mutate(Time.Interval=as.numeric(gsub("Int","",interval)),
         GT.Age=paste(GT,AgeGrp,sep = "_")) %>% 
  pivot_wider(id_cols = c(
    GT,AgeGrp,sample_id,tissueID, mID,
    batch), 
    names_from = interval, values_from = ocr) %>% 
  mutate(Maximal=Int3-Int4,
         Spare = Int3 - Int1,
         Basal = Int1 - Int4,
         ATP_linked = Int1 - Int2,
         Proton_leak = Int2 - Int4,
         Non_Mito = Int4
  ) %>%
  select(-Int1, -Int2, -Int3, -Int4) %>%
  mutate( across(all_of(data.cols),~ mean(.x, na.rm = TRUE)),.by = c(tissueID)) %>%
  select(-sample_id) %>%
  distinct() 

# create in long form for plotting too
dt.states.l <- 
  dt.states %>% 
  pivot_longer(cols=all_of(data.cols),
               names_to = "Measure",
               values_to = "Value") 



```



## Plots

### Respiratory States

```{r fig.height=5, fig.width=7}

ggplot(dt.states.l, aes(AgeGrp,Value,fill=GT))+
  geom_boxplot(position = position_dodge(width = .75))+
  geom_point(position = position_dodge(width = .75))+
  facet_wrap(~Measure,nrow = 2)+
  ylab('OCR (pmol/min)/crypt')+
  xlab("Age")+
  theme_bw()+
  scale_fill_grey()+
  theme(legend.title = element_blank(),
        strip.background = element_rect(fill = "white"),
        legend.position = "right",
        legend.text = element_text(size=11),
        strip.text = element_text(size=10),
  )



```


### Time series

```{r}

dt.norm.means.ocr<-
  dt.norm %>% 
  mutate(tis.time=paste(tissueID,time,sep = "_")) %>% 
  select(c(tis.time,tissueID,time,ocr,
                  GT,AgeGrp)) %>% 
  group_by(tis.time) %>% 
  mutate(ocr=mean(ocr,na.rm=TRUE)) %>% 
  distinct()

datSum.ocr<-summarySEwithin(dt.norm.means.ocr, measurevar="ocr", 
                            withinvars=c("time"), 
                            idvar="tissueID", 
                            betweenvars=c(
                              "GT","AgeGrp"),
                            na.rm=TRUE, conf.interval=.95)

datSum.ocr$time<-as.numeric(as.character(datSum.ocr$time))

dt.norm.means.ecar<-
  dt.norm %>% 
  mutate(tis.time=paste(tissueID,time,sep = "_")) %>% 
  select(c(tis.time,tissueID,time,ecar,
                  GT,AgeGrp)) %>% 
  group_by(tis.time) %>% 
  mutate(ecar=mean(ecar,na.rm=TRUE)) %>% 
  distinct()

datSum.ecar<-summarySEwithin(dt.norm.means.ecar, measurevar="ecar", 
                             withinvars=c("time"), 
                             idvar="tissueID", 
                             betweenvars=c(
                               "GT","AgeGrp"),
                             na.rm=TRUE, conf.interval=.95)

datSum.ecar$time<-as.numeric(as.character(datSum.ecar$time))

```


```{r}

ggplot(datSum.ocr, aes(time, ocr, linetype = GT)) + 
  geom_line() + 
  geom_errorbar(aes(ymin=ocr-se, ymax=ocr+se), colour="black", width=.1, linetype=1) +
  geom_point(size = 1) +
  theme_bw() + xlab('Timepoint') +ylab('OCR (pmol/min)/crypt')+
  scale_x_continuous(breaks = seq(min(datSum.ocr$time), max(datSum.ocr$time),by=1),minor_breaks = waiver())+
  theme(legend.title = element_blank(),
        legend.position = "bottom",
        legend.spacing.y = unit(0,"pt"),
        legend.margin = margin(0,0,0,0),
        plot.margin = margin(5,5,0,5),
        legend.text = element_text(size=12),
        strip.text = element_text(size=11),
        strip.background = element_rect(fill = "white"))+
  facet_grid(~AgeGrp,scales = "free")

```



```{r}

ggplot(datSum.ecar, aes(time, ecar, linetype = GT)) + 
  geom_line() + 
  geom_errorbar(aes(ymin=ecar-se, ymax=ecar+se), colour="black", width=.1, linetype=1) +
  geom_point(size = 1) +
  theme_bw() + xlab('Timepoint') +ylab('ECAR (mpH/min)/crypt')+
  scale_x_continuous(breaks = seq(min(datSum.ecar$time), max(datSum.ecar$time),by=1),minor_breaks = waiver())+
  theme(legend.title = element_blank(),
        legend.position = "bottom",
        legend.spacing.y = unit(0,"pt"),
        legend.margin = margin(0,0,0,0),
        plot.margin = margin(5,5,0,5),
        legend.text = element_text(size=12),
        strip.text = element_text(size=11),
        strip.background = element_rect(fill = "white"))+
  facet_grid(~AgeGrp,scales = "free")

```



## Statistical analysis


### Respiratory states

```{r}

fit.list<-list()

for(v in data.cols){
  fit.list[[v]]<-list()
  mod<-paste(v,"~ GT + AgeGrp + GT:AgeGrp")
  fit.list[[v]][["all"]]<-lm(eval(parse(text=mod)),
                             data=dt.states,
                             na.action="na.exclude")
  
  mod<-paste(v,"~ GT")
  fit.list[[v]][["young"]]<-lm(eval(parse(text=mod)),
                               data=dt.states[dt.states$AgeGrp=="1-4m",],
                               na.action="na.exclude")
  
  fit.list[[v]][["adult"]]<-lm(eval(parse(text=mod)),
                               data=dt.states[dt.states$AgeGrp=="10-12m",],
                               na.action="na.exclude")
}

sum.list<-lapply(fit.list, function(x) lapply(x,summary))

modsig<-do.call(rbind,
                do.call(rbind,
                        lapply(
                          fit.list, function(x) 
                            lapply(x,broom::glance)
                        )))

modsig$age<-rep(names(fit.list$basal_resp),length(fit.list))

Measure<-NULL
for(i in 1:length(fit.list)){
  Measure<-c(Measure,rep(names(fit.list)[i],3))
}

modsig$Measure<-Measure

restab<-NULL
for(v in names(fit.list)){
  for(agegrp in names(fit.list[[v]])){
    sumtab<-summary(fit.list[[v]][[agegrp]])$coefficients
    df<-cbind(
      Measure=rep(v,nrow(sumtab)),
      age=rep(agegrp,nrow(sumtab)),
      term=rownames(sumtab),
      sumtab
    )
    restab<-rbind(restab,df)
  }
  
}

restab <- restab  %>%  
    as.data.frame(row.names = F) %>% 
    mutate_at(c(4:7), as.numeric)

knitr::kable(restab, digits = 3)
# write.csv(restab,file = "regRes.csv")



```



### Time series

```{r}

gt=factor(dt.norm.means.ocr$GT)
ocr=dt.norm.means.ocr$ocr
age=factor(dt.norm.means.ocr$AgeGrp)
t=dt.norm.means.ocr$time

mod_gam1 = mgcv::gam(ocr~s(t)+gt+age+gt*age)
summary(mod_gam1)
mod_gam2 = mgcv::gam(ocr~s(t,by = gt)+gt+age+gt*age)
summary(mod_gam2)
anova(mod_gam1,mod_gam2,test="Chisq")


```



by age group
```{r}

for(age in unique(dt.norm.means.ocr$AgeGrp)){
  cat("\n\n",age,"\n")
  gt=factor(dt.norm.means.ocr$GT[dt.norm.means.ocr$AgeGrp==age])
  ocr=dt.norm.means.ocr$ocr[dt.norm.means.ocr$AgeGrp==age]
  t=dt.norm.means.ocr$time[dt.norm.means.ocr$AgeGrp==age]
  mod_gam1 = mgcv::gam(ocr~s(t)+gt)
  mod_gam2 = mgcv::gam(ocr~s(t,by = gt)+gt)
  print(summary(mod_gam1))
  print(summary(mod_gam2))
  print(anova(mod_gam1,mod_gam2,test="Chisq"))
}

```

